Semantic Linearity in a Lazy Optimising Compiler

### 2025-01-21 icfp.tex
1 and 2:header

We present Linear Core, the first type system that understands semantic
linearity in the presence of non-strict evaluation, suitable for the Core
intermediate language of the Glasgow Haskell Compiler. We prove Linear Core
guarantees linear resource usage via a type safety argument and prove
multiple optimising transformations preserve linearity in Linear Core while
failing to do so in Core. We have implemented Linear Core as a compiler
plugin to validate the system against linearity-heavy libraries, including
\texttt{linear-base}.%^, in the heart of the compiler.
\end{abstract}

@@

Linear type systems~\cite{cite:linear-logic,cite:barberdill} add expressiveness
to existing type systems by enforcing that certain \emph{resources} (e.g.~a
file handle) are used \emph{exactly once}.
%
In programming languages with a linear type system, not using certain resources
or using them twice is flagged as a type error. For instance, linear types can
statically enforce correct usage of file or socket descriptors, or
that heap-allocated memory is freed exactly once (leaks and double-frees become
type errors), or even guarantee deadlock freedom in channel-based communication
protocols~\cite{10.1007/978-3-642-15375-4_16},
% implement safe
% high-performance language interoperability~\cite{}, 
%guarantee that quantum entangled states are not duplicated~\cite{}
among other correctness properties~\cite{10.1145/3373718.3394765,10.1145/3527313,cite:linearhaskell}.
% handle mutable state safely~\cite{}
%
The following program with a runtime error (allocated memory is freed twice) is
accepted by a C-like type system. Conversely, under the lens of a linear type
system where $p$ is deemed a linear resource created by the call to
\texttt{malloc}, the program is rejected:
% We know this to be the dreaded double-free error which will crash the program at runtime.
% \centering
\begin{code}
let p = malloc(4); in free(p); free(p);
\end{code}

@@

Both Haskell and its intermediate language Core are lazily
        evaluated.
        % -- expressions in Haskell are only evaluated when needed.
        As we make clear in this paper, \emph{lazy evaluation interacts non-trivially with linearity}.
        Intuitively, a syntactic occurrence of a linear resource does not
        necessarily entail consuming that resource because expressions are not
        necessarily evaluated (i.e., if the expression is not evaluated, the
        resource is not used).
        %
        Under eager evaluation, a syntactic occurrence of a resource always implies its use at runtime.
        %
        % Under eager/call-by-value evaluation, the distinction between syntactic uses of a
        % resource and the actual use of linear resources at runtime does not exist --
        % an occurrence of a variable in the program always entails consuming it.
        %

@@

Just as Core's type system provides a degree of validation to the
translation from Haskell (dubbed \emph{desugaring}) and the subsequent
optimising transformations, a \emph{linearly typed Core} would guarantee that
linear resource usage in the source language is not violated by desugaring
and the compiler optimisation passes. Moreover, linearity information in
Core can inform Core-to-Core optimising 
transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
such as inlining and $\beta$-reduction, to produce more efficient programs.
%
% It is crucial that a program's behaviour enforced by linear types is \emph{not}
% changed by the compiler in the desugaring or optimisation stages (optimisations
% should not destroy linearity!) and a linearity aware Core type-checker is key in
% providing such guarantees -- it would be disastrous if the compiler, e.g.,
% duplicated a pointer to heap-allocated memory that was previously just used
% once and then freed in the original program.
%TODO: linearidade pode informar otimizacoes
%
% Even more, linearity in Core can inform Core-to-Core optimising
% transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
% including inlining and $\beta$-reduction, to produce more performant programs.

@@

Despite the advantages of a linearly typed Core, the status quo is
that linearity is effectively ignored in Core, despite being checked
in the source Haskell code. The reason is that the desugaring
and the subsequent Core-to-Core optimising transformations
eliminate and rearrange most of the syntactic constructs through
which linearity checking is performed -- often resulting in programs
that are syntactically very different from the original, especially
due to the flexibility laziness provides a compiler in the
optimisations it may perform.
%
Crucially, while optimisations aim to preserve the semantics of
programs and therefore a program's linear resource consumption,
a \emph{naive} syntactic analysis of the optimised programs (as
performed by the current Core type checker) often fails to recognize
the program as linear.
%
For instance, let $x$ be a linear resource in the two following programs, where
the latter results from inlining $y$ in the let body of the former. Even though
the program no longer appears linear (as there are now two 
occurrences of $x$), the program is indeed linear
because the let-bound expression is never evaluated, so $x$ is
consumed exactly once (when it is freed in the let body):
% executed, $x$ will be freed exactly once:
%
% , \emph{is} linear, as the let bound expression freeing $x$
% is never evaluated (because is not needed) -- thus $x$ is only consumed once
% when freed in the let body:
\[
\begin{array}{ccc}
\llet{y = \textsf{free}~x}{y} & \Longrightarrow_{Inlining} & \llet{y = \textsf{free}~x}{\textsf{free}~x}
\end{array}
\]

@@

The Core optimising transformations expose a fundamental limitation of Core's
linear type system: it does not account for the lazy evaluation model
of Core and, thus, a whole class of programs that are linear under the lens of
lazy evaluation are rejected by Core's current linear type system.
%
In this work, we address this limitation by exploring how syntactic linearity
breaks down in the presence of laziness and designing a type system which,
derived from our observations, accepts the vast majority of valid non-strict
programs and optimising transformations previously rejected by traditional
linear typing disciplines. Additionally, we show our system
%
% and its implications on validating and influencing optimising transformations
% by designing a type system which analyses how syntactic linearity breaks down
% in the presence of laziness
is suitable for the intermediate language of an
optimising compiler by implementing it as a Glasgow Haskell Compiler plugin.
%
Our contributions are as follows:
%
\begin{itemize}

@@

\item We introduce Linear Core, a type system for a linear lazy language with
the key features from Core (except for type equality coercions), which,
accounts for semantic linearity in the presence of laziness. To the
best of our knowledge, this is the first type system to capture linearity
semantically in this context
(Section~\ref{sec:main:linear-core}).

\item We show the type soundness of Linear Core and prove that
  multiple optimising transformations, which
fail to type check in Core, are validated by Linear
Core (Section~\ref{sec:main:metatheory}).

@@

A linear type system statically guarantees that linear resources are
\emph{consumed} exactly once. Consequently, whether a program is
well-typed under a linear type system intrinsically depends on the
definition of \emph{consuming} a resource.
In general, \emph{consuming} a resource is equated with its 
\emph{syntactic occurrence} in the program. However, as this section
makes clear, in linear type systems for non-strict
languages, a distinction between using
resources \emph{syntactically} and \emph{semantically} is needed.
%
For instance, in the following Haskell-like program, there are two syntactic
occurrences of \lstinline{handle}: one is in a let-bound computation that
closes this handle, the other is returned. If the program is evaluated lazily,
the handle will not be closed because the computation that closes the handle is
never evaluated (\lstinline{x} goes unused).
%
\begin{lstlisting}
f : Handle *' $\lolli$ '* Handle
f handle = let x = close handle in handle
\end{lstlisting}
%
% Intuitively, a computation that depends on a linear resource to produce a
% result consumes that resource iff the result is effectively computed; in
% contrast, a computation that depends on a linear resource, but is never run,
% will not consume that resource.
%
% Even though there are two occurrences of \lstinline{handle} in the program, and
% it looks as though the \lstinline{handle} may be closed before returned, this function may
% actually only \emph{consume} used once -- it depends on the evaluation strategy of the
% language -- the \lstinline{handle} is closed iff the let bound computation is
% evaluated.
%
%
% Intuitively, \emph{consuming} a resource is not necessarily synonymous with its
% syntactic occurrence. In the example, \emph{non-strict} evaluation means the
% handle is never closed by the function.
%
%
% Depending on the evaluation strategy of the language, a computation that closes
% a handle might not be evaluated, and thus is not consumed.
% With \emph{eager} evaluation semantics,
% the let
%   bound expression \lstinline{close handle} is eagerly evaluated, and the \lstinline{handle} will be closed before being
% returned. It is clear that a linear type system should not accept such a
% program since the linear resource \lstinline{handle} is duplicated -- it is used in a
% computation that closes it, while still being made available to the caller of
% the function;
% %
% on the other hand, with \emph{call-by-need},
%
% Reconsider the \lstinline{handle} example program under
% \emph{call-by-need} evaluation semantics:
% The let bound expression is only evaluated if the binding \lstinline{x} is
% needed. The function returns the \lstinline{handle} right away, and forgets the
% let bound \lstinline{x} as its scope is exited, such that \lstinline{x} will
% never be needed -- consequently, the \lstinline{handle} is never closed by the
% function...
%
%
%

Intuitively, a computation that depends on a linear resource to produce a
result consumes that resource iff the result is effectively consumed; in
contrast, a computation that depends on a linear resource, but is never run,
will not consume that resource.
Under the lens of non-strict evaluation, this means that a syntactic
occurrence of a resource in a let binding only entails consuming that resource
iff the binding itself is consumed.
%
We argue that a linear type system for a language with non-strict evaluation
semantics should accept the above program, unlike a linear type system for the
same program if it were to be evaluated eagerly.
%
From this observation, and exploring the connection between computation and evaluation,
% it becomes clear that \emph{linearity} and \emph{consuming resources}, in the
% above example and for programs in general, should be defined in function of the
% language's evaluation strategy.
%
we turn our focus to linearity under non-strict evaluation, both since GHC Core is
non-strict and because the distinction between semantically and syntactically
consuming a resource is only exposed under non-strict semantics.
%
% Indeed, under \emph{call-by-value}, syntactic occurrences of a linear resource
% directly correspond to semantically using that resource\footnote{With the minor exception
% of trivial aliases, which don't entail any computation even in
% \emph{call-by-value}. In theory, we could use in mutual exclusion any of the
% aliases to refer to a resource without loss of linearity} because \emph{all}
% expressions are eagerly evaluated -- if all computations are eagerly run, all
% linear resources required by computations are \emph{eagerly consumed}.

>>>>

We present Linear Core, the first type system that understands semantic
linearity in the presence of non-strict evaluation, suitable for the Core
intermediate language of the Glasgow Haskell Compiler. We prove Linear Core
guarantees linear resource usage via a type safety and bisimilarity argument and prove
multiple optimising transformations preserve linearity in Linear Core while
failing to do so in Core. We have implemented Linear Core as a compiler
plugin to validate the system against linearity-heavy libraries, including
\texttt{linear-base}.%^, in the heart of the compiler.
\end{abstract}

@@

Linear type systems~\cite{cite:linear-logic,cite:barberdill} add expressiveness
to existing type systems by enforcing that certain resources (e.g.~a
file handle) are used \emph{exactly once}.
%
In programming languages with a linear type system, not using certain resources
or using them twice is flagged as a type error. For instance, linear types can
statically enforce correct usage of file or socket descriptors, or
that heap-allocated memory is freed exactly once (leaks and double-frees become
type errors), or even guarantee deadlock freedom in channel-based communication
protocols~\cite{10.1007/978-3-642-15375-4_16},
% implement safe
% high-performance language interoperability~\cite{}, 
%guarantee that quantum entangled states are not duplicated~\cite{}
among other correctness properties~\cite{10.1145/3373718.3394765,10.1145/3527313,cite:linearhaskell}.
% handle mutable state safely~\cite{}
%
The following program with a runtime error (allocated memory is freed twice) is
accepted by a C-like type system. Conversely, under the lens of a linear type
system where $p$ is deemed a linear resource created by the call to
\texttt{malloc}, the program is rejected:
% We know this to be the dreaded double-free error which will crash the program at runtime.
% \centering
\begin{code}
let p = malloc(4); in free(p); free(p);
\end{code}

@@

Both Haskell and its intermediate language Core are lazily
        evaluated.
        % -- expressions in Haskell are only evaluated when needed.
        As we make clear in this paper, \emph{non-strict evaluation interacts non-trivially with linearity}.
        Intuitively, a syntactic occurrence of a linear resource does not
        necessarily entail consuming that resource because expressions are not
        necessarily evaluated (i.e., if the expression is not evaluated, the
        resource is not used).
        %
        Under eager evaluation, a syntactic occurrence of a resource always implies its use at runtime.
        %
        % Under eager/call-by-value evaluation, the distinction between syntactic uses of a
        % resource and the actual use of linear resources at runtime does not exist --
        % an occurrence of a variable in the program always entails consuming it.
        %

@@

Just as Core's type system provides a degree of validation to the
translation from Haskell (dubbed \emph{desugaring}) and the subsequent
optimising transformations, a linearly typed Core would guarantee that
linear resource usage in the source language is not violated by desugaring
and the compiler optimisation passes. Moreover, linearity information in
Core can inform Core-to-Core optimising 
transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
such as inlining and $\beta$-reduction, to produce more efficient programs.
%
% It is crucial that a program's behaviour enforced by linear types is \emph{not}
% changed by the compiler in the desugaring or optimisation stages (optimisations
% should not destroy linearity!) and a linearity aware Core type-checker is key in
% providing such guarantees -- it would be disastrous if the compiler, e.g.,
% duplicated a pointer to heap-allocated memory that was previously just used
% once and then freed in the original program.
%TODO: linearidade pode informar otimizacoes
%
% Even more, linearity in Core can inform Core-to-Core optimising
% transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
% including inlining and $\beta$-reduction, to produce more performant programs.

@@

Despite the advantages of a linearly typed Core, the status quo is
that linearity is effectively ignored in Core, despite being checked
in the source Haskell code. The reason is that the desugaring
and subsequent Core-to-Core optimising transformations
eliminate and rearrange most of the syntactic constructs through
which linearity checking is performed -- often resulting in programs
that are syntactically very different from the original, especially
due to the flexibility laziness provides a compiler in the
optimisations it may perform.
%
Crucially, while optimisations aim to preserve the semantics of
programs and therefore a program's linear resource consumption,
a \emph{naive} syntactic analysis of the optimised programs (as
performed by the current Core type checker) often fails to recognize
the program as linear.
%
As a trivial example, let $x$ be a linear resource in the next two expressions,
where the latter results from inlining $y$ without eliminating the binder. Even
though the second expression no longer appears linear (as there are now two occurrences
of $x$), it is indeed linear because the let-bound expression is never
evaluated, thus $x$ is still consumed exactly once (when it is freed in the let body):
% executed, $x$ will be freed exactly once:
%
% , \emph{is} linear, as the let bound expression freeing $x$
% is never evaluated (because is not needed) -- thus $x$ is only consumed once
% when freed in the let body:
\[
\begin{array}{ccc}
\llet{y = \textsf{free}~x}{y} & \Longrightarrow_{Inlining} & \llet{y = \textsf{free}~x}{\textsf{free}~x}
\end{array}
\]

@@

The Core optimising transformations expose a fundamental limitation of Core's
linear type system: it does not account for the non-strict evaluation model
of Core and, thus, a whole class of programs that are linear under the lens of
non-strict evaluation are rejected by Core's current linear type system.
%
In this work, we address this limitation by exploring how syntactic linearity
breaks down in the presence of non-strictness and designing a type system which,
derived from our observations, accepts the vast majority of valid non-strict
programs and optimising transformations previously rejected by traditional
linear typing disciplines. Additionally, we show our system
%
% and its implications on validating and influencing optimising transformations
% by designing a type system which analyses how syntactic linearity breaks down
% in the presence of laziness
is suitable for the intermediate language of an
optimising compiler by implementing it as a Glasgow Haskell Compiler plugin.
%
Our contributions are as follows:
%
\begin{itemize}

@@

\item We introduce Linear Core, a non-strict linear language with
the key features from Core (except for type equality coercions), whose type system
accounts for semantic linearity in the presence of laziness. To the
best of our knowledge, this is the first type system to capture linearity
semantically in this context
(Section~\ref{sec:main:linear-core}).

\item We prove our type system is sound, guarantees linear resource usage, and
    prove that multiple optimising transformations, which are rejected in Core,
        are validated by Linear Core (Section~\ref{sec:main:metatheory}).

@@

A linear type system statically guarantees that linear resources are
consumed exactly once. Consequently, whether a program is
well-typed under a linear type system intrinsically depends on the
definition of consuming a resource.
In general, consuming a resource is equated with its 
\emph{syntactic occurrence} in the program. However, as this section
makes clear, in linear type systems for non-strict
languages, a distinction between using
resources \emph{syntactically} and \emph{semantically} is needed.
%
For instance, in the following Haskell-like program, there are two syntactic
occurrences of \lstinline{handle}: one is in a let-bound computation that
closes this handle, the other is returned. If the program is evaluated lazily,
the handle will not be closed because the computation that closes the handle is
never evaluated (\lstinline{x} goes unused).
%
\begin{lstlisting}
f : Handle *' $\lolli$ '* Handle
f handle = let x = close handle in handle
\end{lstlisting}
%
% Intuitively, a computation that depends on a linear resource to produce a
% result consumes that resource iff the result is effectively computed; in
% contrast, a computation that depends on a linear resource, but is never run,
% will not consume that resource.
%
% Even though there are two occurrences of \lstinline{handle} in the program, and
% it looks as though the \lstinline{handle} may be closed before returned, this function may
% actually only \emph{consume} used once -- it depends on the evaluation strategy of the
% language -- the \lstinline{handle} is closed iff the let bound computation is
% evaluated.
%
%
% Intuitively, \emph{consuming} a resource is not necessarily synonymous with its
% syntactic occurrence. In the example, \emph{non-strict} evaluation means the
% handle is never closed by the function.
%
%
% Depending on the evaluation strategy of the language, a computation that closes
% a handle might not be evaluated, and thus is not consumed.
% With \emph{eager} evaluation semantics,
% the let
%   bound expression \lstinline{close handle} is eagerly evaluated, and the \lstinline{handle} will be closed before being
% returned. It is clear that a linear type system should not accept such a
% program since the linear resource \lstinline{handle} is duplicated -- it is used in a
% computation that closes it, while still being made available to the caller of
% the function;
% %
% on the other hand, with \emph{call-by-need},
%
% Reconsider the \lstinline{handle} example program under
% \emph{call-by-need} evaluation semantics:
% The let bound expression is only evaluated if the binding \lstinline{x} is
% needed. The function returns the \lstinline{handle} right away, and forgets the
% let bound \lstinline{x} as its scope is exited, such that \lstinline{x} will
% never be needed -- consequently, the \lstinline{handle} is never closed by the
% function...
%
%
%

%Under the lens of non-strict evaluation, a syntactic
%occurrence of a resource in an expression only entails consuming that resource
%iff the expression itself is fully consumed.
%
Intuitively, a computation that depends on a linear resource to produce a
result consumes that resource iff the result is fully consumed; in
contrast, a computation that depends on a linear resource, but is never run,
will not consume that resource.
%
We argue that a linear type system for a language with non-strict evaluation
semantics should accept the above program, unlike a linear type system for the
same program if it were to be evaluated eagerly.
%
From this observation, and exploring the connection between computation and evaluation,
% it becomes clear that \emph{linearity} and \emph{consuming resources}, in the
% above example and for programs in general, should be defined in function of the
% language's evaluation strategy.
%
we turn our focus to linearity under non-strict evaluation, both since GHC Core is
lazy and because the distinction between semantically and syntactically
consuming a resource is only exposed under non-strict semantics.
%
% Indeed, under \emph{call-by-value}, syntactic occurrences of a linear resource
% directly correspond to semantically using that resource\footnote{With the minor exception
% of trivial aliases, which don't entail any computation even in
% \emph{call-by-value}. In theory, we could use in mutual exclusion any of the
% aliases to refer to a resource without loss of linearity} because \emph{all}
% expressions are eagerly evaluated -- if all computations are eagerly run, all
% linear resources required by computations are \emph{eagerly consumed}.
